---
title: "Practical Machine Learning Course Project"
output:
  html_document:
    keep_md: yes
date: "Saturday, January 14, 2017"
---

```{r setup, echo=FALSE}
echo = TRUE                               # Always make code visible
options(scipen = 1)                       # Turn off scientific notations for numbers
knitr::opts_chunk$set(cache=FALSE)         # Caching is on by default
```

## Background
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. This project uses data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset). The goal of the project is to predict the manner in which they did the exercise (the "classe" variable in the training set).

## Load Libraries
```{r results='hide',message=FALSE,warning=FALSE}
library(caret)
library(randomForest)
library(ggplot2)
```

## Load Raw Training and Test Data
```{r}
training.file <- "pml-training.csv"
testing.file  <- "pml-testing.csv"

# Check if the data files exist locally, if not retrieve
if ( !file.exists(training.file) ) {
  download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv", destfile=training.file)  
}
if ( !file.exists(testing.file) ) {
  download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv", destfile=testing.file)  
}

# Load data with minimal cleanup
training.raw <- read.csv(training.file, na.strings = c("NA", "#DIV/0!", ""))
testing.raw  <- read.csv(testing.file,  na.strings = c("NA", "#DIV/0!", ""))
```

## Cleaning the Data
```{r}
dim(training.raw)       # Size of training data
dim(testing.raw)        # Size of testing data
```

Check for variables with more than 50% NAs:
```{r}
NA_vars <- sapply(training.raw, function(x) {sum(is.na(x))}) >  0.5*dim(training.raw)[1]
```

The initial training data set is of size `r dim(training.raw)[1]` observations by `r dim(training.raw)[2]` variables. However there are `r sum(NA_vars)` variables with 50% or more of the observations missing. Interpolating such a large number of missing observations by imputing values seems unsatisfactory. Therefore we will exclude these variables:
```{r}
training.clean <- training.raw[,!NA_vars]
dim(training.clean)                      # Size of partly cleaned non-NA training data
```

Next, check for and remove any variables which have zero or near zero variance predictors:
```{r}
NZP <- nearZeroVar(training.clean,saveMetrics=TRUE)
training.clean <- training.clean[,!NZP$nzv]
dim(training.clean)                     # Size of partly cleaned training data having removed NZ predictors
```

Examine the surviving variables
```{r}
str(training.clean)
```

Sanity check to see if any NAs left in the data set
```{r}
anyNA(training.clean)                  # No NAs left in training data set
```

The first variable "X" appears to be an index of the observations. The variable "user_name" seems to an id of the participant. The variables "raw_timestamp_part_1", "raw_timestamp_part_2" and "cvtd_timestamp" all seem to be date/time stamps. The variable "num_window" appears to be another index variable, perhaps for the group of observations within a particular time window for each participant. As these variables seem to be secondary to the recorded variables of interest they will therefore be removed from the subsequent analysis.
```{r}
training.clean <- training.clean[,-c(1:6)]
dim(training.clean)                    # Final Size of cleaned training data
```

Preprocess the data to center and scale it
```{r}
preObj <- preProcess(training.clean[,-53], method=c("center","scale"))
training.clean.pp <- data.frame(cbind(classe=training.clean[,53],predict(preObj, training.clean[,-53])))
```

Apply same series of cleanup steps to the testing data:
```{r}
testing.clean <- testing.raw[,!NA_vars]
testing.clean <- testing.clean[,!NZP$nzv]
testing.clean <- testing.clean[,-c(1:6)]
testing.clean.pp <- data.frame(cbind(classe=testing.clean[,53],predict(preObj, testing.clean[,-53])))
dim(testing.clean.pp)                 # Final Size of cleaned testing data (same number of vars as training.clean)
```

## Split the data
Having cleaned the data, now split it into a subset for training (70% of cases) and an independent subset for data validation (remaining 30% of cases):
```{r}
set.seed(8421)                         # Set seed to a particular value to ensure reproducibility
inTrain   <- createDataPartition(training.clean.pp$classe, p=0.70, list=F)
train <- training.clean.pp[inTrain, ]
valid <- training.clean.pp[-inTrain, ]
```

## Model Building
LDA, Random Forest and GBM models were all run and tested. LDA and GBM models were significantly lesss accurate than the Random Forest model. Therefore the Random Forest model was the preferred modelling choice. For brevity, only the final Random Forest analysis is shown here. A Random Forest model was run on the cleaned up training data. 10-fold cross validation was used to create an in-sample error estimate of the error.

## Fit Random Forest model with 10-fold cross validation
```{r}
fitCtrl <- trainControl(method = "cv", number=10)
# Select a range of mtry parameters to try to fit over
grid <- expand.grid(mtry=c(5,7,10,12,15,20,25,30))
grid

# This model fit takes a long time to run with default ntree=500. 
# Check if the model has already been saved and if so load it, otherwise calculate afresh
if (file.exists("rf_model.rda")) {
  load("rf_model.rda")
} else {
 rf_model <- train(classe ~ ., data = train, method = "rf", trControl = fitCtrl, tuneGrid=grid)
 save(rf_model, file="rf_model.rda")
}

# OOB Error (as a percentage)
OOB_error <- round(100*(1-sum(diag(rf_model$finalModel$confusion))/dim(train)[1]),2)
OOB_error
```

## Fitted model
```{r}
# Characteristics of fitted model
rf_model
rf_model$finalModel

# Show the relative importance of top 30 variables
plot(varImp(rf_model),30)

# Show the relationship between the top 3 variables
# There is some separability between the classes but also some overlap
qplot(roll_belt, yaw_belt, color=classe, data=train)
qplot(roll_belt, pitch_forearm, color=classe, data=train)
qplot(yaw_belt, pitch_forearm, color=classe, data=train)
```

## Out-of-sample error
```{r}
# Use the independent validation data to test the out of sample error via the prediction accuracy 
valid_pred <- predict(rf_model, valid)
confusionMatrix(valid_pred, valid$classe)

# Out-of-sample error rate (as a percentage)
OOS_error <- round(100*(1-mean(valid_pred == valid$classe)),2)
OOS_error
```

## Conclusion
From the bivariate plots separation of classes might be expected to be quite difficult. However it can be seen that the Random Forest model provides a very high degree of classification accuracy. The out-of-bag (OOB) error estimated via the 10-fold cross validation within the training sample was only `r OOB_error`%. However, the unbiased out-of-sample error rate estimated from the independent sample was actually slightly smaller at `r OOS_error`%. 

## Predictions on the hold-out Testing data set
```{r}
pred_test <- predict(rf_model, testing.clean.pp)
pred_test
```